---
title: "Project X"
---

Welcome to our project page.

Keep an eye on this space to stay updated with our project activities.


---
title: "Population Movements After the February 6, 2023 Earthquake"
authors: Büşra Halil and Şevval Miray Başar
date: 06.04.2026
format: html
code-fold: true
---

# 1. Project Overview and Scope

The February 6, 2023 Kahramanmaraş earthquakes, with magnitudes 
of 7.7 and 7.6, are among the deadliest natural disasters in 
Turkey's modern history. The disaster caused widespread destruction 
across 11 provinces, displacing hundreds of thousands of people 
and triggering one of the largest internal migration movements 
the country has ever seen.

This project analyzes the demographic impact of the earthquakes 
by examining population changes in the affected provinces before 
and after the disaster. Using TurkStat's Address-Based Population 
Registration System (ADNKS), the study investigates how the 
earthquakes reshaped population distribution across Turkey.

**Research questions:**

- How did the populations of the 11 affected provinces change 
  between 2022 and 2023?
- Which provinces experienced the greatest population loss?
- Is there a relationship between earthquake damage intensity 
  and population decline?

**Objectives:**

- Quantify the population loss in earthquake-affected provinces
- Identify migration patterns following the disaster
- Visualize the demographic impact through maps and charts

# 2. Data
## Data

This project uses two complementary datasets from TurkStat (TÜİK):

**1. Address-Based Population Registration System (ADNKS)**  
Provincial population figures for the 11 earthquake-affected 
provinces across 2018–2025, used to quantify overall population 
change before and after the disaster.

**2. Inter-Provincial Migration Statistics**  
The primary dataset for this analysis. It records migration flows 
between all province pairs where at least one province is among 
the 11 earthquake-affected provinces, covering 2018–2025. Each 
row represents a directed province pair for a given year and 
includes the following variables:

| Variable | Description |
|---|---|
| `Yıl` | Year (2018–2025) |
| `Göç Alan İl` | Receiving province |
| `Göç Veren İl` | Sending province |
| `Göç Alan İl Nüfus` | Population of receiving province |
| `Göç Veren İl Nüfus` | Population of sending province |
| `Aldığı Göç` | In-migration count |
| `Verdiği Göç` | Out-migration count |
| `Net Göç` | Net migration (in minus out) |

The dataset contains 13,200 observations. The 2023 records 
capture migration flows immediately following the February 6 
earthquakes, enabling a direct before-and-after comparison of 
migration patterns.

**Sources:** TurkStat ADNKS & Inter-Provincial Migration 
Statistics — [data.tuik.gov.tr](https://data.tuik.gov.tr)  
**RData file:** [deprem_goc.RData](your_github_link_here)
```{r}
#| message: false
#| warning: false

library(readxl)
library(tidyverse)

# Population data
df_nufus <- read_excel("tr_nufus.xlsx")

# Migration data
df_goc <- read_excel("deprem_goc_ham_filtreli.xlsx", skip = 2)

save(df_nufus, df_goc, file = "deprem_goc.RData")

glimpse(df_goc)
```

## 2.2 General Information About Data

The dataset covers the 11 provinces directly affected by the 
February 6, 2023 earthquakes: Kahramanmaraş, Hatay, Adıyaman, 
Malatya, Gaziantep, Şanlıurfa, Diyarbakır, Kilis, Osmaniye, 
Adana, and Elazığ.

Each observation corresponds to one province and includes the 
following variables:

| Variable | Description | Type |
|---|---|---|
| `province` | Province name | Character |
| `population_22` | Population before earthquake (2022) | Numeric |
| `population_23` | Population after earthquake (2023) | Numeric |
| `growth_rate` | Annual population growth rate (‰) | Numeric |
| `pop_change` | Population change (persons) | Numeric |


## 2.3 Reason of Choice

This dataset  is publicly available, regularly updated, and directly published 
by TurkStat, ensuring reliability. Also, the 2022 and 2023 
editions capture the population status immediately before and 
after the earthquakes, making it ideal for measuring the 
demographic impact. Additionally, the provincial level of granularity 
is well-suited for comparative analysis across the affected region.

The February 6 earthquakes represent a critical turning point 
in Turkey's demographic landscape. Studying their impact on 
population movement contributes to our understanding of how 
large-scale disasters reshape human geography.

## 2.4 Preprocessing

The raw data was downloaded from TurkStat in Excel format. 
The following preprocessing steps were applied:

1. **Column selection:** Only the province name, population and annual growth rate columns were retained.
2. **Province filtering:** The dataset was filtered to include 
   only the 11 earthquake-affected provinces.
3. **Type conversion:** Population figures contained spaces as 
   thousand separators and were converted to numeric format.
4. **Derived variable:** A population change variable (`pop_change`) 
   was computed as the difference between 2023 and 2022 populations.
```{r}
#| message: false
#| warning: false
#| label: data-preprocessing

library(readxl)
library(tidyverse)

earthquake_provinces <- c("Kahramanmaraş", "Hatay", "Adıyaman",
                          "Malatya", "Gaziantep", "Şanlıurfa",
                          "Diyarbakır", "Kilis", "Osmaniye",
                          "Adana", "Elazığ")

df_raw <- read_excel("adnks_il.xls", skip = 2)

df <- df_raw |>
  rename(
    province      = 1,
    population_22 = 2,
    population_23 = 3,
    growth_rate   = 10
  ) |>
  select(province, population_22, population_23, growth_rate) |>
  filter(province %in% earthquake_provinces) |>
  mutate(
    population_22 = as.numeric(gsub(" ", "", population_22)),
    population_23 = as.numeric(gsub(" ", "", population_23)),
    growth_rate   = as.numeric(growth_rate),
    pop_change    = population_23 - population_22
  )

save(df, file = "deprem_goc.RData")
```
```{r}
#| label: data-overview
#| tbl-cap: "Population change in earthquake-affected provinces (2022–2023)"

df |>
  arrange(growth_rate) |>
  knitr::kable(
    col.names = c("Province", "Population 2022", "Population 2023",
                  "Growth Rate (‰)", "Change (persons)"),
    format.args = list(big.mark = ",")
  )
df <- read_excel("tr_nufus.xlsx")

save(df, file = "deprem_goc.RData")

glimpse(df)
```
```




